---
title: "gene_set_analysis"
author: "Carlos"
date: "2025-01-22"
output: 
  html_document:
    toc: yes
    toc_float:
      smooth_scroll: yes
---

::: questions
-   What is the aim of performing gene set enrichment analysis?
-   What is the method of over-representation analysis?
-   What are the commonly-used gene set databases?
:::

```{r load-pkg, eval = FALSE, results='hide', message=F, warning=F}
library(SummarizedExperiment)
library(DESeq2)
library(gplots)
library(microbenchmark)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(msigdbr)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(simplifyEnrichment)
library(dplyr)
library(fgsea)
library(tibble)
library(topGO)
```

Load the data (again)

```{r error = TRUE, eval=F, results='hide', message=F, warning=F}
library(SummarizedExperiment)
library(DESeq2)

se <- readRDS("/home/carlosaugustodmm/biocrnaseq_cfbioinfo/GSE96870_se.rds")
se <- se[rowSums(assay(se, "counts")) > 5, ]
```

```{r read-data, warning = FALSE}
library(SummarizedExperiment)
library(DESeq2)

# only restrict to mRNA (protein-coding genes)
se <- se[rowData(se)$gbkey == "mRNA"]
# construct a `DESeqDataSet` object where we also specify the experimental design
dds <- DESeqDataSet(se, design = ~ sex + time)
# perform DESeq2 analysis
dds <- DESeq(dds)
# obtain DESeq2 results, here we only want Male vs Female in the "sex" variable
resSex <- results(dds, contrast = c("sex", "Male", "Female"))
# extract DE genes with padj < 0.05
sexDE <- as.data.frame(subset(resSex, padj < 0.05))
# the list of DE genes
sexDEgenes <- rownames(sexDE)
```

```{r xy-genes}
geneGR <- rowRanges(se) #`GRanges` object
totalGenes <- rownames(se)
XYGeneSet <- totalGenes[as.vector(seqnames(geneGR)) %in% c("X", "Y")]
head(XYGeneSet)
length(XYGeneSet)
```

Visualize intersection with Venn diagram

```{r venn-diagram, fig.width = 5*2, fig.height = 5*2, dpi = 72*2}
library(gplots)
venn(list("sexDEgenes"  = sexDEgenes, 
               "XY gene set" = XYGeneSet))
```

### Fisher's exact test

2x2 contingency table:

|            | In the gene set | Not in the gene set | Total    |
|------------|-----------------|---------------------|----------|
| **DE**     | $n_{11}$        | $n_{12}$            | $n_{1+}$ |
| **Not DE** | $n_{21}$        | $n_{22}$            | $n_{2+}$ |
| **Total**  | $n_{+1}$        | $n_{+2}$            | $n$      |

\*Note we replace `+` with `0` in the R variable names. Genes must be
unique in each vector.

```{r fisher-1}
n    <- nrow(se)
n_01 <- length(XYGeneSet)
n_10 <- length(sexDEgenes)
n_11 <- length(intersect(sexDEgenes, XYGeneSet))
```

Other values can be obtained by:

```{r fisher-2}
n_12 <- n_10 - n_11
n_21 <- n_01 - n_11
n_20 <- n    - n_10
n_02 <- n    - n_01
n_22 <- n_02 - n_12
```

All the values are:

```{r fisher-3}
matrix(c(n_11, n_12, n_10, n_21, n_22, n_20, n_01, n_02, n),
    nrow = 3, byrow = TRUE)
```

Is there a dependency of a gene to be a DE gene and to be in the *XY
gene set*?

```{r fisher-test}
t <- fisher.test(matrix(c(n_11, n_12, n_21, n_22), nrow = 2, byrow = TRUE),
    alternative = "greater")
t$p.value
```

In the output, we can see the *p*-value is very small, then we can
conclude DE genes have a very strong enrichment in the *XY gene set*.

## Gene set resources

#### topGO 
```{r shrinking}
res_8_vs_0_shrunk <- lfcShrink(dds,
                               coef="time_Day8_vs_Day0", type="apeglm") 
timeDE_shrunk <- as.data.frame(res_8_vs_0_shrunk)
```

```{r rank and find ids}
# Ranking genes by log2FC:
rankedGenes <- timeDE_shrunk$log2FoldChange
names(rankedGenes) <- rownames(timeDE_shrunk)

# Extract symbols from rankedGenes and map
geneSymbols <- names(rankedGenes)
entrezIDs <- mapIds(org.Mm.eg.db,
  keys = geneSymbols,
  column = "ENTREZID",
  keytype = "SYMBOL",
  multiVals = "first")
names(rankedGenes) <- entrezIDs
rankedGenes <- rankedGenes[!is.na(names(rankedGenes))] # removes NAs

# GO-to-EntrezIds mapping
GO_2_genes <- as.list(org.Mm.egGO2ALLEGS) # get GO-to-EntrezIds correspondence
```

```{r create topGO object and summarize}
# topGOdata object
GOdata <- new("topGOdata",
  ontology = "BP",
  allGenes = rankedGenes,
  geneSelectionFun = function(x) abs(x) > 1.5,
  annot = annFUN.GO2genes,
  GO2genes = GO_2_genes)

# run fisher
resultFisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")

topGOtable <- GenTable(GOdata, classicFisher = resultFisher, topNodes = length(resultFisher@score))
```

Plotting
```{r topGO plot (bad example)}
topGOtable$classicFisher <- as.numeric(topGOtable$classicFisher) # because otherwise it's set as `chr`
topGOtable$EnrichmentRatio <- topGOtable$Significant / topGOtable$Expected

topGOtable_top <- topGOtable %>%
  as_tibble() %>%
  filter(classicFisher < 0.05) %>%
  arrange(classicFisher) %>%
  slice_head(n = 24)

ggplot(topGOtable_top, aes(x= -log10(classicFisher), y=(reorder(Term, -log10(classicFisher))))) +
  geom_col(aes(fill=EnrichmentRatio)) +
  labs(x="-log10(p-value)", y="Pathway") +
  theme_minimal()
```

#### MSigDB gene sets

-   Manually curated gene set database
-   Collections can be seen at https://www.gsea-msigdb.org/gsea/msigdb/index.jsp

Check which organisms are supported and which gene sets collections it
provides:

```{r msigdbr}
library(msigdbr)
msigdbr_species()
msigdbr_collections()
```

#### fgsea

```{r downloading msigdbr C2 gene set}
library(fgsea)
library(msigdbr)
library(tibble)
library(dplyr)

# testing with shrunk results
# get C2 gene set:
gene_sets_C2 <- msigdbr(category = "C2", species = "mouse")
head(gene_sets_C2)

#setting it as a list format to use fgsea on:
gene_sets_C2_list <- gene_sets_C2 %>%
  group_by(gs_name) %>%
  summarize(genes = list(gene_symbol)) %>%
  deframe() #converts 2 column df to a named list
```

-   Shrinking the results and using `fgsea()`

```{r shrink and enrichment}
#repeating the results:
res_8_vs_0_shrunk <- lfcShrink(dds,
                               coef="time_Day8_vs_Day0", type="apeglm") 
timeDE_shrunk <- as.data.frame(res_8_vs_0_shrunk)

timeDE_shrunk_rank <- timeDE_shrunk$log2FoldChange #creating a vector
names(timeDE_shrunk_rank) <- rownames(timeDE_shrunk) #rownames
head(timeDE_shrunk_rank)

fgseaRes <- fgsea(pathways = gene_sets_C2_list,
               stats = timeDE_shrunk_rank,
               maxSize = 400)

fgseaRes %>% #rank top 10 |NES|
    arrange(desc(abs(NES)))
```

-   Plotting the enrichment

```{r plot}
fgseaRes_top_point <- fgseaRes %>%
  as_tibble() %>%
  filter(padj < 0.05) %>%
  arrange(desc(NES)) %>%
  slice_head(n = 12) %>%
  bind_rows(fgseaRes %>% # to arrange both top up and down regulated
      as_tibble() %>%
      filter(padj < 0.05) %>%
      arrange(NES) %>%
      slice_head(n = 12))

ggplot(fgseaRes_top_point, aes(y=(reorder(pathway, NES)), x=0)) +
  geom_point(aes(fill = NES, size = size), shape = 21) +
  scale_fill_gradient2(low = "#327eba", high = "#e06663", midpoint = 0) +
  labs(y="Pathway", x=NULL)+
  theme_minimal() +
  scale_x_continuous(breaks = NULL)
```
