---
title: "Exploratory_analysis"
author: "Iza"
date: "2025-01-17"
output: html_document
---
Suggested bibliography:
original bioc-rnaseq:https://carpentries-incubator.github.io/bioc-rnaseq/

## 4. Exploratory analysis and quality control
Questions:
* Why is exploratory analysis an essential part of an RNA-seq analysis?
* How should one preprocess the raw count matrix for exploratory analysis?
* Are two dimensions sufficient to represent your data?

Load packages
```{r}
suppressPackageStartupMessages({
    library(SummarizedExperiment)
    library(DESeq2)
    library(vsn)
    library(ggplot2)
    library(ComplexHeatmap)
    library(RColorBrewer)
    library(hexbin)
    library(iSEE)
})
```

Read your SE
```{r}
se <- readRDS("data/GSE96870_se.rds")

nrow(se)
```

Remove genes with less than 5 coutns
```{r}
# Remove genes/rows that do not have > 5 total counts 
se <- se[rowSums(assay(se, "counts")) > 5, ]
nrow(se)
```

Differences in library sizes
```{r}
# Add in the sum of all counts

se$libSize <-  colSums(assay(se))

# Plot the libSize by using R's native pipe |>
# to extract the colData, turn it into a regular
# data frame then send to ggplot:

colData(se) |>
  as.data.frame() |>
  ggplot(aes(x = Label, y = libSize / 1e6, fill = Group)) + 
         geom_bar(stat = "identity") + theme_bw() + 
         labs(x = "Sample", y = "Total count in millions") + 
         theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

```

Estimate size factors using Deseq2
```{r}
dds <- DESeq2::DESeqDataSet(se, design = ~ sex + time)

dds <- estimateSizeFactors(dds)

# Plot the size factors against library size
# and look for any patterns by group:

ggplot(data.frame(libSize = colSums(assay(dds)),
                  sizeFactor = sizeFactors(dds),
                  Group = dds$Group),
       aes(x = libSize, y = sizeFactor, col = Group)) + 
    geom_point(size = 5) + theme_bw() + 
    labs(x = "Library size", y = "Size factor")
```

Transforming the data
```{r}
meanSdPlot(assay(dds), ranks = FALSE)
```

```{r}
vsd <- DESeq2::vst(dds, blind = TRUE)
meanSdPlot(assay(vsd), ranks = FALSE)

```

Heatmaps and clustering
```{r}
dst <- dist(t(assay(vsd)))
colors <- colorRampPalette(brewer.pal(9, "Blues"))(255)
ComplexHeatmap::Heatmap(
    as.matrix(dst), 
    col = colors,
    name = "Euclidean\ndistance",
    cluster_rows = hclust(dst),
    cluster_columns = hclust(dst),
    bottom_annotation = columnAnnotation(
        sex = vsd$sex,
        time = vsd$time,
        col = list(sex = c(Female = "red", Male = "blue"),
                   time = c(Day0 = "yellow", Day4 = "forestgreen", Day8 = "purple")))
)
```


PCA

###Iza: Use the Wine analogy to explain PCA?

```{r}
pcaData <- DESeq2::plotPCA(vsd, intgroup = c("sex", "time"),
                           returnData = TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(x = PC1, y = PC2)) +
    geom_point(aes(color = sex, shape = time), size = 5) +
    theme_minimal() +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    coord_fixed() + 
    scale_color_manual(values = c(Male = "blue", Female = "red"))
```

Interactive exploratory with iSEE
```{r}
## Convert DESeqDataSet object to a SingleCellExperiment object, in order to 
## be able to store the PCA representation
sce <- as(dds, "SingleCellExperiment")

## Add PCA to the 'reducedDim' slot
stopifnot(rownames(pcaData) == colnames(sce))
reducedDim(sce, "PCA") <- as.matrix(pcaData[, c("PC1", "PC2")])

## Add variance-stabilized data as a new assay
stopifnot(colnames(vsd) == colnames(sce))
assay(sce, "vsd") <- assay(vsd)

app <- iSEE(sce)
shiny::runApp(app)

```



